
###Load packages
```{r packages, echo=T, message=F, warnings=F}
library(lme4) # mixed effect models
library(ggplot2) # graphics
library(dplyr) # data manipulation

```

###Load data
```{r setup}
setwd("~/Downloads")
deer.data<-read.csv('deer data for selection analysis.csv',header=T,stringsAsFactors = T)  # read in the data

```


### Visualising the data


```{r visualise}
head(deer.data)
```

```{r factors}
deer.data$Sex<-as.factor(deer.data$Sex)
deer.data$BirthYear<-as.factor(deer.data$BirthYear)
```

Summarise the data in the column Birth Weight:
```{r describe birth weight, echo=F, message=F, warnings=F}
summary(deer.data$BirthWt)

#hist(deer.data$BirthWt.tr)
ggplot(deer.data, aes(BirthWt)) + 
  geom_histogram(binwidth = .2,colour="black",fill="white")+
  xlab("Birth weight (kg)")+
  theme_bw()
```

### Distribution of birth weight split by sex
```{r split by sex, message=F, warnings=F}
tapply(deer.data$BirthWt, deer.data$Sex, summary)

ggplot(deer.data, aes(BirthWt, fill = Sex)) + 
  geom_histogram(binwidth = .2,position="dodge")+
  xlab("Birth weight (kg)")+
  theme_bw()
```

Alternatively looking at the variation in birth weights of the two sexes across the years:
```{r split by year, message=F, warnings=F}

ggplot(deer.data, aes(BirthYear, BirthWt, col=Sex))+ 
  geom_point()+ 
  xlab("Year")+
  ylab("Calf birth weight (kg)")+
  theme(axis.text.x = element_text(angle = -90, hjust=0))+  # can't get axis labels to flip
  theme_bw()
```

... which shows a lot of variation in the data




### Fit a linear model of birth weight 


```{r linear model of birth weight, message=F, warnings=F}
lm1<-lm(BirthWt ~ Sex, data=deer.data)
summary(lm1)
```
Very low R-squared value reflects the large amount of variation in the data


## Fit a linear mixed model of birth weight
The linear model above does not take into account the structuring in the data, with calves born in different years and to different mothers. Here I used a mixed effects model

```{r linear mixed effects model, message=F, warnings=F}
lme1<-lmer(BirthWt ~ Sex + (1|BirthYear) + (1|Mum), data=deer.data)
summary(lme1)
```
Confirms the statistical difference between the sexes

## Check diagnostics of the model of birth weight
``` {r check residuals, message=F, warnings=F}
plot(lme1)
qqnorm(residuals(lme1))
qqline(residuals(lme1))
```

Sufficient for this purpose.

# Models of selection
We will now look at the natural selection acting on birth weight, by quantifying its association with components of fitness.

The data file contains columns with Lifetime Breeding Success (LBS) for males and females separately: this is the total number of offspring that an individual produces in its lifetime.
The distribution of these values:
```{r summarise LBS values}
summary(deer.data$FemaleLBS)  # summarise female LBS
ggplot(deer.data, aes(FemaleLBS)) + 
  geom_histogram(binwidth = 1,colour="black",fill="lightgrey")+
  xlab('Female Lifetime Breeding Success')+
  theme_bw()
  
summary(deer.data$MaleLBS) # summarise male LBS
ggplot(deer.data, aes(MaleLBS)) + 
  geom_histogram(binwidth = 1,colour="black",fill="darkgrey")+
  xlab('Male Lifetime Breeding Success')+
  theme_bw()
```

Lifetime breeding success is a difficult distribution to analyse, so we will therefore split it into two components: 

survival to adulthood ('juvenile survival'), 
and then breeding success amongst those that survive to adulthood ('adult breeding success'). 

### Natural selection via variation in juvenile survival
We define 'juvenile survival' by whether or not an individual lives to age 3.
```{r define juvenile survival and adult breeding success}
deer.data <- deer.data %>%
  mutate(JuvSurv=ifelse(AgeAtDeath>2,1,0)) 

# quick summary table
table(deer.data$JuvSurv,deer.data$AgeAtDeath)
```

### Does juvenile survival depend on birth weight?

```{r glm}
tapply(deer.data$BirthWt, deer.data$JuvSurv,mean)
```

### Estimating a 'selection differential' for birth weight via juvenile survival

The 'selection differential' on a trait is the difference in the mean before vs after selection.
It is the difference between the mean birth weight of all calves, compared to the mean of those that survived.

```{r difference in means}

mean(deer.data$BirthWt)
mean(deer.data$BirthWt[deer.data$JuvSurv==1],na.rm=T)
mean(deer.data$BirthWt[deer.data$JuvSurv==1],na.rm=T) - mean(deer.data$BirthWt,na.rm=T)

```
Another way to estimate the selection differential on a trait is the statistical covariance between the trait and relative fitness, where relative fitness is the fitness measure standardised to a mean of 1. 


```{r covariance}
deer.data <- deer.data %>% 
  mutate(rel_JuvSurv=JuvSurv/mean(JuvSurv))   # creates a new column of relative juvenile survival
CovJS<-cov(deer.data$rel_JuvSurv,deer.data$BirthWt,use="pairwise.complete.obs")    # estimates covariance with birth weight
CovJS
```


```{r rescaling to get population covariance}
N<-length(deer.data$BirthWt)   # number of observations
CovJS * (N-1)/N
```
### Estimating a 'selection gradient' for selection on birth weight via juvenile survival
This is the gradient (slope) of a linear regression of relative fitness on the trait.

```{r selection gradient via juvenile survival}
lm.juvsurv<-lm(rel_JuvSurv ~ Sex+BirthWt, data=deer.data)
summary(lm.juvsurv)

```


# Statistical models of selection on birth weight 
Using a GLMM this time.

## Selection model for juvenile survival, with both sexes together
```{r glmm juvenile survival}
glmm.juvsurv<-glmer(JuvSurv ~ Sex+BirthWt+(1|Mum)+(1|BirthYear), data=deer.data, family='binomial')
summary(glmm.juvsurv)
```




## Selection via adult breeding success
Now considering selection via adult breeding success
```{r define adult LBS, message=F, warnings=F}
deer.data<-deer.data %>%
  mutate(Fem_AdultLBS=ifelse(AgeAtDeath>2,FemaleLBS,NA)) %>%  
  mutate(Male_AdultLBS=ifelse(AgeAtDeath>2,MaleLBS,NA))        

ggplot(deer.data, aes(Fem_AdultLBS)) + 
  geom_histogram(binwidth = 1,colour="black",fill="lightgrey")+
  theme_bw()+ 
  xlab("Female Adult LBS")

ggplot(deer.data, aes(Male_AdultLBS)) + 
  geom_histogram(binwidth = 1,colour="black",fill="darkgrey")+
  theme_bw()+ 
  xlab("Male Adult LBS")
```



Female Adult LBS follows a Poisson distribution, Male Adult LBS is much more skewed and needs a Negative Binomial.

### Females:  selection on birth weight via adult LBS
Plotting the data to look at the relationship between birth weight and adult LBS. 
```{r plot female adult LBS birthwt}
ggplot(deer.data, aes(BirthWt,Fem_AdultLBS))+ 
  geom_point()+ 
  xlab("birth weight (kg)")+
  ylab("Female Adult LBS")+
  theme(axis.text.x = element_text(angle = -90, hjust=0))+  # can't get axis labels to flip
  theme_bw()
```
GLMM for adult lifetime breeding success in females, specifying a Poisson distribution

``` {r glmm female adult LBS}
glmm.Fem_AdultLBS<-glmer(Fem_AdultLBS ~ BirthWt+(1|Mum)+(1|BirthYear), data=deer.data, family='poisson')
summary(glmm.Fem_AdultLBS)
```

## Males:  selection model for adult LBS

```{r glmm.nb male adult LBS}
ggplot(deer.data, aes(BirthWt,Male_AdultLBS))+ 
  geom_point()+ 
  xlab("birth weight (kg)")+
  ylab("Male Adult LBS")+
  theme(axis.text.x = element_text(angle = -90, hjust=0))+  # can't get axis labels to flip
  theme_bw()

glmm.Male_AdultLBS<-glmer.nb(Male_AdultLBS ~ BirthWt+(1|BirthYear), data=deer.data)
summary(glmm.Male_AdultLBS)
```

### Estimating selection differentials and selection gradients for birth weight via adult breeding success
Estimates of the two selection coefficients in either sex via adult lifetime breeding success. First, defining the relative fitness  for LBS in each sex.
```{r define relative LBS}
deer.data<-deer.data %>%
  mutate(relFem_AdultLBS=Fem_AdultLBS/mean(Fem_AdultLBS,na.rm=T)) %>%
  mutate(relMale_AdultLBS=Male_AdultLBS/mean(Male_AdultLBS,na.rm=T))    
# check that it's done it right
mean(deer.data$relFem_AdultLBS,na.rm=T)
mean(deer.data$relMale_AdultLBS,na.rm=T)
```

#### Selection differentials via adult LBS
For adult breeding success, I estimate a selection differential in either sex from the covariance with relative LBS.
```{r selection gradients for adult LBS}
cov(deer.data$BirthWt,deer.data$relFem_AdultLBS,use="complete.obs") 
cov(deer.data$BirthWt,deer.data$relMale_AdultLBS,use="complete.obs")
```

#### Selection gradients
Finally, selection gradients* can be taken from linear regressions of relative fitness in either sex.
```{r selection differentials for adult LBS}
lm.femLBS<-lm(relFem_AdultLBS~BirthWt,data=deer.data) 
summary(lm.femLBS)$coeff[2,1:2]
lm.maleLBS<-lm(relMale_AdultLBS~BirthWt,data=deer.data) 
summary(lm.maleLBS)$coeff[2,1:2] # female and male selection gradients are printed

```